---
title: "Tidy Tuesday #5"
description: |
  Tidy Tuesday 10/26/21
author:
  - name: Yunyang Zhong
date: 10-26-2021
preview: tidy.PNG
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=FALSE}
library(tidyverse)
library(lubridate)
library(ggThemeAssist)

ultra_rankings <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-10-26/ultra_rankings.csv')
race <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-10-26/race.csv')
```

```{r}
ultra_rankings %>% 
  inner_join(race, by = 'race_year_id') %>%
  filter(race_year_id == 54013) %>%
  filter(!is.na(gender)) %>% 
  mutate(res = hms(time),
         hour = hour(res) + minute(res)/60 + second(res)/60^2) %>% 
  group_by(age, gender) %>% 
  mutate(speed = distance/mean(hour)) %>% 
  ggplot(aes(x = age, y = speed, color = gender)) +
  geom_line() +
  theme(plot.title = element_text(size = 15)) +
  labs(title="What is the average speed (km/h) for each gender and age to \nfinish P'tiot Sparnatrail?", caption = "Plot created by Yunyang Zhong,\ndata from Benjamin Nowak by way of International Trail Running Association (ITRA)") +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5), panel.background = element_rect(fill = "gray90"))
```

> I created this graph to help answer what is the average speed (km/h) for each gender and age to finish P'tiot Sparnatrail. I first filtered the data to only P'tiot Sparnatrail and then created a new variable representing hour using lubridate. After grouped by gender and age, average speed was computed using distance and average hour. The title was made bigger and the y-axis-title, speed, was made horizontal to make it easier to read. It seems that 30-year-old men and 37-year-old women are the fastest while 72-year-old men and 28-year-old women the slowest.
